Assignment 1: Course Project Reproducible Research
==================================================

Assignment 1 requires the following packages to be installed:

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library("knitr")
opts_chunk$set(echo = TRUE)
```

Reading Data from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%Factivity.zip

```{r}
  setwd("C:/Users/Owner/Documents")
  data <- read.csv("activity.csv")
```

Clean the data a bit: 

```{r} 
  data$date <- ymd(data$date)
```

Clean data check: 

```{r}
  str(data)
  head(data)
```

Data is now clean. Now we can begin to answer the questions that have put forward in this assignment. 

Question 1: What is the mean total number of steps taken per day? 

  1. Calculate the total number of steps taken/day
  
  2. If you do not understand the difference between a histogram and a bar plot, research the difference.
     A good start could be: http://stattrek.com/statistics/charts/histogram.apsx?Tutorial=AP
     
  3. Calculate and report the mean and median of the total number of steps taken each day.
  
```{r}
steps <- data %>%
  filter(!is.na(steps))%>%
  group_by(date)%>%
  summarize(steps = sum(steps))%>%
  print()
```

Now we can use ggplot to make the histogram: 

```{r} 
ggplot(steps, aes(x =steps)) +
  geom_histogram(fill = "blue", binwidth = 1000) +
  labs(title = "Histogram of Steps/Day", x = "Steps/Day", y = "Frequency")
```

Last part of question 1 requires us to get the mean & median of Steps/Day

```{r} 
mean_steps <- mean(steps$steps, na.rm = TRUE)
median_steps <- median(steps$steps, na.rm = TRUE)
```


Question 2: What is the average daily activity pattern? 

1. What a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps
   taken, averaged across all days (y-axis)?

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of
   steps?

```{r}
interval <- data %>% 
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))

ggplot(interval, aes(x = interval, y = steps)) +
  geom_line(color = "blue")
```

Now use [which.max] to establish the maximum steps (on average) across all these days. 

```{r}
interval[which.max(interval$steps),]
```

Question 3: Imputing missing values

 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 
    NAs).

 2. Devise a strategy for filling in all of the missing values in the dataset. 
    The strategy does not need to be sophisticated. For example, you could use the mean/median for that 
    day, or the mean for that 5-minute interval, etc.

 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

 4. Make a histogram of the total number of steps taken each day and calculate and report the mean 
    and median total number of steps taken per day. Do these values differ from the estimates from the 
    first part of the assignment? What is the impact of imputing missing data on the estimates of the 
    total daily number of steps?

First lets summarize all the missing values: 
```{r}
sum(is.na(data$steps))
```

Now we can work on we can work on resolving the NA values & create a new data set.
```{r}
data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm = TRUE, simplify = TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

That should have removed all NA values, but lets make sure. 
```{r}
sum(is.na(data_full$steps))
```

If above finished correctly then you should recieve a number in return of zero, if so, then we can
make a summary.

```{r}
steps_full <- data_full %>%
  filter(!is.na(steps))%>%
  group_by(date)%>%
  summarize(steps = sum(steps))%>%
  print
```

Then use ggplot to create the histogram needed

```{r}
ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "Blue", binwidth = 1000) +
  labs(title = "Histogram of Steps/Day (Including Missing Values)", x = "Steps/Day", y = "Frequency")
```
 
Now we can calculate the mean/median steps with the filled in values

```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
median_steps_full <- median(steps_full$steps, na.rm = TRUE)
```

Question 4: Are there differences in activity patterns between weekdays & weekends?

 1. Create a new factor variable in the dataset with two levels - ???weekday??? and ???weekend??? indicating 
    whether a given date is a weekday or weekend day.

 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
    and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


First we need to create a new column for weektype. This is done through using [dplyr] and [mutate].

```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)
```

Now we can calculate the average steps in the 5-minute interval. Lets use ggplot to compare the average
steps for weeday and weekend. 

```{r}
interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```